---
title: "Analysis of Public Health Data"
subtitle: 'Introduction to R and Data Management'
author: ""
date: "23/06/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Textbook 
•	Kirkwood B and Sterne J. Essential Medical Statistics, 2nd Edition. May 2003, ©2003, Wiley-Blackwell. 

#### References 
•	Donegan K, King B and Bryan P. Safety of pertussis vaccination in pregnant women in UK: observational study. BMJ 2014;349:g4219 doi: 10.1136/bmj.g4219 (Published 11 July 2014). 

•	Gregoire AJ, Kumar R, Everitt B, Henderson AF, Studd JW. Transdermal oestrogen for treatment of severe postnatal depression. Lancet. 1996 Apr 6;347(9006):930-3. 

•	Rabe-Hesketh S and Everitt B. A Handbook of Statistical Analyses Using R; Fourth Edition. 2007, Chapman & Hall/CRC. 

•	Spanos A, Harrell FE, Durack DT. Differential diagnosis of acute meningitis: An analysis of the predictive value of initial observations. JAMA, 1989; 262: 2700-2707. 

#### Datasets 

##### *Meningitis data* 
+ Meningitis csv dataset: dataset in csv (comma delimited) format 

##### *Pertussis data*
+ pertussis.csv 

##### *Dates data*
+ date exercise.csv: date exercise in csv (comma delimited) format 

##### *Postnatal depression data* 
+	depress1a.csv
+ depress1b.csv 
+ depress2.csv 
+ depress3b.csv 
+ depress4_5.csv 
+ depress6a.csv 
+ depress6b.csv 

## Module Objectives 
This module provides an introduction to the R statistical package that you will be using for POPH8303/POPH8913 Analysis of Public Health Data. On completion of this module you should:


1. Have an understanding of basic features of the R statistical program 
2. Be able to undertake basic R commands to input, view, describe and save data 
3. Be able to generate basic graphs 
4. Be familiar with the R environment 
5. Have an understanding of the requirements for data checking and data cleaning 
6. Be able to undertake manipulation of variables, including dates 
7. Be able to undertake manipulation of datasets, such as appending, merging and reshaping


## Introduction to R and Describing Data

R is a very powerful, and relatively easy to use Statistical package which allows you to undertake data management and manipulation, perform a range of basic and complex statistical tests and analyses, and generate a wide variety of different graphs. For this course, you will be required to use R to undertake basic analysis and graphics.

R is a dialect of the S language. It is a case-sensitive, interpreted language. You can enter commands one at a time at the command prompt (>) or run a set of commands from a source file. There is a wide variety of data types, including vectors (numerical, character, logical), matrices, data frames, and lists. Most functionality is provided through built-in and user-created functions and all data objects are kept in memory during an interactive session. Basic functions are available by default. Other functions are contained in packages that can be attached to a current session as needed.

R packages are bundles of functions which extend the capabilities of R. Thousands of add-on packages are available in the main online repository (CRAN) and several more packages can be found on GitHub. These packages can be found and installed online. 

Stata and R have minor difference in default setting and methods. In this document, we will follow the Stata analysis as closely as possible, but small and usually important difference in Stata and R will be noted. Unlike Stata, R can hold multiple datasets in memory simultaneously and so, there is no need to save intermediate files or close and re-open datasets. 

You need to communicate with R in a way that the program understands; i.e. you need to be able to tell R what you want in a way that it understands. You can do this by typing in commands using an appropriate “language”. You should save a copy of your commands, often referred to as “program”, so that you can check and repeat the analysis and so that you have a permanent record of what you have done.

Syntax, also sometimes referred to as “code” or “programming”, refers to the grammatical rules that you need to follow so that R understands what you want it to do. Syntax does not necessarily follow rules of English grammar! You tell R to do something by using the correct command or set of commands. An R command is a combination of syntax elements, expressed following the rules of ‘R language’. 


## R Studio

R Studio is an integrated development environment (IDE) for R. It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, and debugging and workspace management.  It is available in open source and commercial editions on the desktop (Windows, Mac, and Linux) and from a web browser to a Linux server running R Studio Server or R Studio Server Pro. In this course, we will be using R Studio. The R Studio environment is similar to the Stata environment and we hope will help ease the transition from Stata to R. 

## The R Program 



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
